Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.

### 2022-12-28 paper_comparison/paper_comparison/paper_comparison.Rmd
fix issue - Package xcolor Error: Undefined color `RedOrange'

---
title: |
  Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a
    email: emi.tanaka@monash.edu
  - name: Susan VanderPlas
    affil: b
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: paper.bib
abstract: |
  Abstract to fill.
keywords: |
  data visualization; visual inference; hypothesis testing; residual plots;
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \newcommand*{\Perm}[2]{{}^{#1}\!P_{#2}}%
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

Experiment I is designed to study the ability of human subjects to detect the effect of a non-linear term $\boldsymbol{z}$ constructed using Hermite polynomials [@hermite1864nouveau] on random vector $\boldsymbol{x}$ formulated as

@@

### Subject allocation
\newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}

@@

@@

@@

>>>>

---
title: |
  Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a
    email: emi.tanaka@monash.edu
  - name: Susan VanderPlas
    affil: b
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: paper.bib
abstract: |
  Abstract to fill.
keywords: |
  data visualization; visual inference; hypothesis testing; residual plots;
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{xcolor}
  \definecolor{orange-red}{rgb}{1.0, 0.27, 0.0}
  \newcommand{\svp}[1]{{\textcolor{orange-red}{#1}}}
  \newcommand*{\Perm}[2]{{}^{#1}\!P_{#2}}%
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

(include English reference)

Experiment I is designed to study the ability of human subjects to detect the effect of a non-linear term $\boldsymbol{z}$ constructed using Hermite polynomials on random vector $\boldsymbol{x}$ formulated as

@@

### Subject allocation

@@

(remove human perception)

(revisit Emi's concern)

@@

(1. How you calc rejection, 2. scatter plot p-value vs p-value -> appendix, 3. percentage in explained the tex, both marginal and conditional, 4. interesting lineup, discuss the reason)

@@

```{r eval = FALSE}
vi_survey %>%
  filter(!null_lineup) %>%
  filter(!attention_check) %>%
  mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
  left_join(vi_survey %>% 
           count(unique_lineup_id) %>%
           rename(num_subject = n)) %>%
  group_by(unique_lineup_id) %>%
  summarise(across(everything(), first)) %>%
  mutate(type = factor(type, levels = c("non-linearity", "heteroskedasticity"))) %>%
  mutate(conv_reject = ifelse(conventional_p_value < 0.05, "Reject", "Not"),
         reject = ifelse(p_value < 0.05, "Reject", "Not")) %>%
  filter(type == "non-linearity", conv_reject == "Not", reject == "Reject") %>%
  View()
```

### 2022-12-28 paper_comparison/paper_comparison/paper_comparison.tex
add mosaic plot
fix issue - Package xcolor Error: Undefined color `RedOrange'

\usepackage{lscape}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\newcommand*{\Perm}[2]{{}^{#1}\!P_{#2}}%

@@

Experiment I is designed to study the ability of human subjects to
detect the effect of a non-linear term \(\boldsymbol{z}\) constructed
using Hermite polynomials \citep{hermite1864nouveau} on random vector
\(\boldsymbol{x}\) formulated as

@@

As mentioned in Section \ref{calculating-p-values-for-the-visual-test},
\(\alpha\) used in Equation \ref{eq:pvalue-beta-binomial} needs to be
estimated using Rorschach lineups. Hence, 36 lineups with all
combinations of \(n\) and \(X_{raw}\) and three replications were
included in experiment III. In these lineups, the data of the data plot
was generated from a model consistent with the null hypothesis, while
the data of the 19 null plots were generated using the same simulation
method discussed in Section \ref{sampling-from-the-null-distribution}.
Technically, this was an alternative to Rorschach lineup but with the
null distribution further conditioning on the estimated coefficients and
the residual sum of squares. Thus, we used another valid method
suggested in \citet{vanderplas2021statistical} which was designed for
typical lineups to estimate \(\alpha\). Considering the differnece
between typical lineups and Rorschach lineups, we have also included a
sensitivity analysis in the Appendix to examine the impact of the
variance of the estimate on our findings.

@@

(discuss choice of tests, mention pkg \texttt{lmtest},
\texttt{skedastic})

@@

To compare the power of visual test and conventional tests, we picked
RESET test (\texttt{resettest}) and Breusch-Pagan test (\texttt{bptest})
from the R package \texttt{lmtest}, and Shapiro-Wilk test
(\texttt{shapiro.test}) from the built-in R package \texttt{stats}.
These three tests can be used to detect non-linearity,
heteroskedasticity and non-normality respectively. Their estimated power
is shown in Figure \ref{fig:polypower}. For the RESET test, we included
different power of fitted values as proxies. According to
\citep{ramsey_tests_1969}, there is no general rule for the power of the
fitted values needed by the RESET test, but it found power up to four is
usually sufficient. Thus, we followed this guideline to conduct the
RESET test. For the Breusch-Pagan test, both \(\boldsymbol{x}\) and
\(\boldsymbol{x}^2\) were included in the auxiliary regression equation
to {[}justify this choice?{]}.

@@

For the non-linearity model, the power of all four tests increases as
the effect gets larger. The power curve of RESET test climbs
aggressively from 13\% to around 70\% as \(log_e(E)\) increases from 0
to 2, while power of other tests respond inactively to the change of
effect and remain lower than 25\% throughout the period, showing that
F-test is way more sensitive to the type of model defects that being
considered. Meanwhile, no noticeable visual features can be spotted from
the example residuals plots.

@@

Overall, the power analysis suggests that conventional tests do not
match how humans perceive residual plots and thus determine model
violations in these two scenarios. Non-severe model violations with
human acceptable residual departures could often alarm by the
conventional tests. This could limit the popularity of conventional
tests in residual diagnostics among analysts.

@@

@@

\caption{Visual test p-value compared to conventional test p-value for lineups produced by heteroskedasticity and non-linearity model. The scatter plot is drawn on a square root scale. Every dot on the plot corresponds to a lineup. Most of the dots are fallen on the left of the y = x line indicating visual test p-value is almost always higher than conventional test p-value. The blue curve is a smoothing of the dots showing a positive trend. }\label{fig:p-value-comparison}
\end{figure}

>>>>

\usepackage{lscape}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{xcolor}
\definecolor{orange-red}{rgb}{1.0, 0.27, 0.0}
\newcommand{\svp}[1]{{\textcolor{orange-red}{#1}}}
\newcommand*{\Perm}[2]{{}^{#1}\!P_{#2}}%

@@

(include English reference)

Experiment I is designed to study the ability of human subjects to
detect the effect of a non-linear term \(\boldsymbol{z}\) constructed
using Hermite polynomials on random vector \(\boldsymbol{x}\) formulated
as

@@

As mentioned in Section \ref{calculating-p-values-for-the-visual-test},
\(\alpha\) used in Equation \ref{eq:pvalue-beta-binomial} needs to be
estimated using {\textcolor{orange-red}{null}} lineups. Hence, 36
lineups with all combinations of \(n\) and \(X_{raw}\) and three
replications were included in experiment III. In these lineups, the data
of the data plot was generated from a model
{\textcolor{orange-red}{with zero effect size}}, while the data of the
19 null plots were generated using the same simulation method discussed
in Section \ref{sampling-from-the-null-distribution}.
{\textcolor{orange-red}{This generation procedure differs from the canonical Rorschach lineup procedure, which requires that all 20 plots are generated from the null hypothesis. However, these lineups serve the same fundamental purpose: to assess the number of visually interesting plots generated under the null hypothesis.}}
{\textcolor{orange-red}{To account for the fact that our simulation method for these lineups is not the Rorschach procedure,}}
we used the method suggested in \citet{vanderplas2021statistical} for
typical lineups containing a data plot to estimate \(\alpha\). We have
included a sensitivity analysis in the Appendix to examine the impact of
the variance of the \(\alpha\) estimate on our findings.

@@

@@

To compare the power of visual test and conventional test, we picked
RESET test (\texttt{resettest}) and Breusch-Pagan test (\texttt{bptest})
from the R package \texttt{lmtest}, and Shapiro-Wilk test
(\texttt{shapiro.test}) from the built-in R package \texttt{stats}.
Among them, RESET test is the only exact and appropriate test in this
scenario. Both the Breusch-Pagan test and the Shapiro-Wilk test are
approximate and inappropriate tests. Their estimated power is shown in
Figure \ref{fig:polypower}. For the RESET test, we included different
power of fitted values as proxies. According to
\citet{ramsey_tests_1969}, there is no general rule for the power of the
fitted values needed by the RESET test, but it found power up to four is
usually sufficient. Thus, we followed this guideline to conduct the
RESET test. For the Breusch-Pagan test, both \(\boldsymbol{x}\) and
\(\boldsymbol{x}^2\) were included in the auxiliary regression equation
to {[}justify this choice?{]}.

@@

For the non-linearity model, the power of all four tests increases as
the effect gets larger. The power curve of RESET test climbs
aggressively from 13\% to around 70\% as \(log_e(E)\) increases from 0
to 2, while power of other tests respond inactively to the change of
effect and remain lower than 25\% throughout the period, showing that
REST test is way more sensitive to the type of model defects that being
considered. Meanwhile, no noticeable visual features can be spotted from
the example residuals plots.

@@

(remove human perception)

(revisit Emi's concern)

Overall, the power analysis suggests that conventional tests do not
match how humans perceive residual plots in these two scenarios. Humans
have much higher tolerance of the residual departures than the
conventional test. Since fail to reject \(H_0\) in a visual test means
that there is no obvious visual discoveries found in the residual plot,
analysts and the general public as the consumers of the output may not
be convinced of the existence of significant residual departures in
spite of the rejection of \(H_0\) given by the conventional test. Even
if the rejection is accepted, the model violation may be considered as
impactless due to the fact that they are not clearly visible. Besides,
the sensitivity of the conventional test could also distract and
discourage analysts from finding simple but good linear approximation to
the data. The rejection of \(H_0\) because of human acceptable and
negligible residual departures is not practically meaningful and useful.
(it does not fit the purpose of )

This could limit the popularity of conventional tests in residual
diagnostics among analysts.

@@

(1. How you calc rejection, 2. scatter plot p-value vs p-value
-\textgreater{} appendix, 3. percentage in explained the tex, both
marginal and conditional, 4. interesting lineup, discuss the reason)

@@

\caption{Rejection rate of visual test compared to conventional test on lineups produced by heteroskedasticity and non-linearity model. Based on the rejection decision made by the visual test and the conventional test, the mosaic plot shows the proportion of lineups in each category.}\label{fig:p-value-comparison}
\end{figure}
